Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-25 paper.qmd
feat: small suggestions

---
title: "Nix for Polyglot, Reproducible Data Science Workflows"
format:
    jss-pdf:
        keep-tex: true
    jss-html:
        embed-resources: true
        toc: true
author:
  - name: Bruno Rodrigues
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg
        country: Luxembourg
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co
abstract: |
  Reproducible analysis requires more than clean, well-documented code:
  researchers must also manage software dependencies, computational
  environments, and workflow execution. Existing tools like [Docker]{.pkg} and
  [renv]{.pkg} address parts of this challenge, but no single solution handles
  polyglot environments, system dependencies, and workflow orchestration
  together. I present two [R]{.proglang} packages that leverage [Nix]{.pkg} to
  provide both reproducible polyglot environments and declarative workflow
  orchestration. [rix]{.pkg} generates [Nix]{.pkg} expressions that define
  reproducible environments spanning [R]{.proglang}, [Python]{.proglang},
  [Julia]{.proglang}, and system dependencies. Building on these environments,
  [rixpress]{.pkg} (and its [Python]{.proglang} port, [ryxpress]{.pkg})
  orchestrates polyglot pipelines where each computational step runs in its own
  hermetically sealed environment, with automatic caching and dependency
  tracking. This approach enables researchers to develop analyses interactively
  while maintaining bit-for-bit reproducibility, supports collaboration across
  heterogeneous computational environments, and ensures that analyses remain
  executable years into the future.

@@

By *linked and executable code and data*, Peng referred to compiled source code
and runnable scripts. In this paper, I interpret this notion more broadly as the
*computational environment*: the complete set of software required to execute an
analysis. Here too, a continuum exists. At the minimal end, authors might only
name the main software used—say, the [R]{.proglang} programming language
[@r-core-team]. More careful authors might also specify the version of
[R]{.proglang}, or list the additional packages and their versions. Rarely,
however, do authors specify the operating system on which the analysis was
performed, even though differences in operating systems can lead to divergent
results when using the same code and software versions, as shown by
@neupane2019. It is even less common for authors to provide step-by-step
installation instructions for the required software stack, an omission often
driven by institutional constraints. Journals, for instance, can inadvertently
hinder reproducibility by imposing strict page or word limits that leave no room
for the necessary technical documentation, discouraging thoroughness in the name
of brevity.

@@

This paper focuses on two critical but often overlooked aspects of
reproducibility: *computational environment management* and *workflow
orchestration*. I present a comprehensive framework addressing both challenges
using the [Nix]{.pkg} package manager [@dolstra2004nix], making it accessible to
researchers through two [R]{.proglang} packages: [rix]{.pkg} and
[rixpress]{.pkg}. Before introducing these packages, I survey existing tools and
their limitations to contextualise my contribution.

@@

When an author includes this information, others attempting to reproduce the
study (future readers, collaborators, or even the author at a later time) can
see which version of [R]{.proglang} and which packages (with their versions)
were used. However, reproducing the environment still requires manually
installing the correct package versions—a process that becomes difficult when
packages depend on system-level libraries. For example, the [sf]{.pkg} package
requires [GDAL]{.pkg}, [GEOS]{.pkg}, and [PROJ]{.pkg} libraries. Installing and
configuring these dependencies varies substantially across operating systems and
can be prohibitively difficult on some platforms.

@@

In the [R]{.proglang} ecosystem, virtual environments are not built-in as they
are in [Python]{.proglang}, but tools like [renv]{.pkg} provide equivalent
functionality. [renv]{.pkg} captures the project’s software state and writes it
to a lock file (`renv.lock`), which includes the exact versions of
[R]{.proglang} and all required packages. This lockfile serves as a blueprint
for restoring the environment automatically, ensuring others can recreate the
same setup with minimal effort.

@@

Writing [Nix]{.pkg} expressions can be challenging for users unfamiliar with the
[Nix]{.proglang} language. However, the ability to define a fully reproducible
development environment in a single text file and then rebuild it anywhere is
highly appealing. To lower the barrier to adoption of [Nix]{.pkg} for
reproducibility, I developed the [rix]{.pkg} package.

>>>>

---
title: "Nix for Polyglot, Reproducible Data Science Workflows"
format:
    jss-pdf:
        keep-tex: true
    jss-html:
        embed-resources: true
        toc: true
author:
  - name: Bruno Rodrigues
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg
        country: Luxembourg
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co
  - name: Philipp Baumann
    affiliations:
      - name: Alliance SwissPass, Switzerland
        department: Data and Analytics
        address: 7, Länggassstrasse
        city: Bern
        country: Switzerland
        postal-code: 3012
    orcid: 0000-0002-3194-8975
    email: baumann-philipp@protonmail.com
abstract: |
  Reproducible analysis requires more than clean, well-documented code:
  researchers must also manage software dependencies, computational
  environments, and workflow execution. Existing tools like [Docker]{.pkg} and
  [renv]{.pkg} address parts of this challenge, but no single solution handles
  polyglot environments, system dependencies, and workflow orchestration
  together. We present two [R]{.proglang} packages that leverage [Nix]{.pkg} to
  provide both reproducible polyglot environments and declarative workflow
  orchestration. [rix]{.pkg} generates [Nix]{.pkg} expressions that define
  reproducible environments spanning [R]{.proglang}, [Python]{.proglang},
  [Julia]{.proglang}, and system dependencies. Building on these environments,
  [rixpress]{.pkg} (and its [Python]{.proglang} port, [ryxpress]{.pkg})
  orchestrate polyglot pipelines. Each computational step runs in its own
  hermetically sealed environment, with automatic caching and dependency
  tracking. This approach enables researchers to develop analyses interactively
  while maintaining bit-for-bit reproducibility. Beyond, it supports
  collaboration across heterogeneous computational environments, and ensures
  that analyses can be reproduced years into the future.

@@

By *linked and executable code and data*, Peng referred to compiled source code
and runnable scripts. In this paper, we interpret this notion more broadly as the
*computational environment*: the complete set of software required to execute an
analysis. Here too, a continuum exists. At the minimal end, authors might only
name the main software used—say, the [R]{.proglang} programming language
[@r-core-team]. More careful authors might also specify the version of
[R]{.proglang}, or list the additional packages and their versions. Rarely,
however, do authors specify the operating system on which the analysis was
performed, even though differences in operating systems can lead to divergent
results when using the same code and software versions, as shown by
@neupane2019. It is even less common for authors to provide step-by-step
installation instructions for the required software stack, an omission often
driven by institutional constraints. Journals, for instance, can inadvertently
hinder reproducibility by imposing strict page or word limits that leave no room
for the necessary technical documentation, discouraging thoroughness in the name
of brevity.

@@

This paper focuses on two critical but often overlooked aspects of
reproducibility: *computational environment management* and *workflow
orchestration*. We present a comprehensive framework addressing both challenges
using the [Nix]{.pkg} package manager [@dolstra2004nix], making it accessible to
researchers through two [R]{.proglang} packages: [rix]{.pkg} and
[rixpress]{.pkg}. Before introducing these packages, we survey existing tools and
their limitations to contextualise our contributions.

@@

When an author includes this information, others attempting to reproduce the
study (future readers, collaborators, or even the author at a later time) can
see which version of [R]{.proglang} and which packages (with their versions)
were used. However, reproducing the environment still requires manually
installing the correct package versions --- a process that becomes difficult when
packages depend on system-level libraries. For example, the [sf]{.pkg} package
requires [GDAL]{.pkg}, [GEOS]{.pkg}, and [PROJ]{.pkg} libraries. Installing and
configuring these dependencies varies substantially across operating systems and
can be prohibitively difficult on some platforms.

@@

In the [R]{.proglang} ecosystem, virtual environments are not built-in as they
are in [Python]{.proglang}, however tools like [renv]{.pkg} provide equivalent
functionality. [renv]{.pkg} captures the project’s software state and writes it
to a lock file (`renv.lock`), which includes the exact versions of
[R]{.proglang} and all required packages. This lockfile serves as a blueprint
for restoring the environment automatically, ensuring others can recreate the
same setup with minimal effort.

@@

Writing [Nix]{.pkg} expressions can be challenging for users unfamiliar with the
[Nix]{.proglang} language. However, the ability to define a fully reproducible
development environment in a single text file and then rebuild it anywhere is
highly appealing. [rix]{.pkg} aims to lower the barrier to adoption of [Nix]{.pkg} for
reproducibility.
